### 2024-06-12 code/eda-code/eda2.qmd
Commit changes

---
title: "Exploratory analysis script"
date: "2024-02-07"
output: html_document
---

This Quarto file loads the cleaned data and does some exploring.

I'm only showing it the way where the code is included in the file. 
As described in the `processing_code` materials, I currently prefer the approach of having R code in a separate file and pulling it in.

But I already had this written and haven't yet re-done it that way. Feel free to redo and send a pull request on GitHub :)

Again, it is largely a matter of preference and what makes the most sense to decide if one wants to have code inside Quarto files, or as separate R files.
And sometimes, an R script with enough comments is good enough and one doesn't need a Quarto file.

Also note that while here I split cleaning and exploring, this is iterative. You saw that as part of the processing, we already had to explore the data somewhat to understand how to clean it. In general, as you explore, you'll find things that need cleaning. As you clean, you can explore more. Therefore, at times it might make more sense to combine the cleaning and exploring code parts into a single R or Quarto file. Or split things in any other logical way.

As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. Figures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data.

Start by exploring one variable at a time. Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables.

Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.

# Setup

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
```

Load the data.

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata2.rds")
#load data
mydata <- readRDS(data_location)
```

# Data exploration through tables

Showing a bit of code to produce and save a summary table.

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "tables", "summarytable1.rds")
saveRDS(summary_df, file = summarytable_file)
```

```{r}
summary_data = summary(mydata)
print(summary_data)
# save to file
summarytable_file = here("results", "tables", "summarytable2.rds")
saveRDS(summary_data, file = summarytable_file)
```

We are saving the results to the `results` folder. Depending on how many tables/figures you have, it might make sense to have separate folders for each. And/or you could have separate folders for exploratory tables/figures and for final tables/figures. Just choose a setup that makes sense for your project and works for you, and provide enough documentation that someone can understand what you are doing.

# Data exploration through figures

Find outliers.

BMI first.

```{r}
p1 <- mydata %>% ggplot(aes(x = "", y = bmi)) + geom_boxplot() + labs(title = "BMI Outliers")
plot(p1)
figure_file = here("results", "figures","bmi_outliers.png")
ggsave(filename = figure_file, plot=p1) 
```

Now Charges.

```{r}
p2 <- mydata %>% ggplot(aes(x = "", y = charges)) + geom_boxplot() + labs(title = "Charges Outliers")
plot(p2)
figure_file = here("results", "figures","charges-outliers.png")
ggsave(filename = figure_file, plot=p2) 
```

Now histogram of insurance charges.

```{r}
p3 <- mydata %>% ggplot(aes(x=charges)) +
  geom_histogram(binwidth=2000, fill="blue", color="black", alpha=0.7) +
  labs(title="Distribution of Insurance Charges", x="Charges", y="Frequency") +
  theme_minimal()
plot(p3)
figure_file = here("results", "figures", "Charges-histogram.png")
ggsave(filename = figure_file, plot=p3) 
```

Once more height as function of weight, stratified by gender. Note that there is so little data, it's a bit silly. But we'll plot it anyway.

```{r}
p4 <- mydata %>% ggplot(aes(x=smoker, y=charges, fill=smoker)) +
  geom_boxplot() +
  labs(title="Insurance Charges by Smoking Status", x="Smoker", y="Charges") +
  scale_fill_manual(values=c("yes"="red", "no"="green")) +
  theme_minimal()
plot(p4)
figure_file = here("results", "figures", "charges-smoking-boxplot.png")
ggsave(filename = figure_file, plot=p4) 
```

```{r}
p5 <- mydata %>% ggplot(aes(x=region, y=charges, fill=region)) +
  geom_boxplot() +
  labs(title="Insurance Charges by Region", x="Region", y="Charges") +
  theme_minimal()
plot(p5)
figure_file = here("results", "figures", "charges-region-boxplot.png")
ggsave(filename = figure_file, plot=p5) 
```

```{r}
p6 <- mydata %>% ggplot(aes(x=age, y=charges)) +
  geom_point(alpha=0.5, color="blue") +
  labs(title="Insurance Charges by Age", x="Age", y="Charges") +
  theme_minimal()
plot(p6)
figure_file = here("results", "figures", "charge-age-boxplot.png")
ggsave(filename = figure_file, plot=p6) 
```

```{r}
p7 <- mydata %>% ggplot(aes(x=bmi, y=charges)) +
  geom_point(alpha=0.5, color="green") +
  labs(title="Insurance Charges by BMI", x="BMI", y="Charges") +
  theme_minimal()
plot(p7)
figure_file = here("results", "figures", "charge-bmi-scatter.png")
ggsave(filename = figure_file, plot=p7) 
```

```{r}
p8 <- mydata %>% ggplot(aes(x=as.factor(children), y=charges, fill=as.factor(children))) +
  stat_summary(fun.y=mean, geom="bar") +
  labs(title="Average Insurance Charges by Number of Children", x="Number of Children", y="Average Charges") +
  theme_minimal()
plot(p8)
figure_file = here("results", "figures", "children-bar.png")
ggsave(filename = figure_file, plot=p8) 
```

# Notes

For your own explorations, tables and figures can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible.

### 2024-06-12 products/manuscript/Manuscript.qmd
Commit changes

---
title: "Data Analysis Project"
subtitle: DA-6833
author: Rinu Thomas
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/apa.csl
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(dplyr)
library(ggplot2)
```

# Summary/Abstract

The primary goal of this research project is to develop a predictive model to estimate medical insurance costs for individuals based on their health-related and personal attributes. This model will help in understanding the key factors influencing insurance charges and assist insurance companies in pricing policies more accurately.

# Introduction

## General Background Information

### Importance of Predictive Modeling in Insurance

Predictive modeling plays a vital role in the insurance industry. Accurate predictions of insurance costs enable companies to:

-   Assess risk more effectively.

-   Set premiums that reflect the actual risk associated with individual policyholders.

-   Enhance customer satisfaction by offering personalized and fair pricing.

On the consumer side, individuals can benefit from understanding the factors that influence their insurance costs, allowing them to make informed choices about their health and financial planning.

## Description of data and data source

*Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section.*

The dataset used for this analysis is sourced from Kaggle. The dataset comprises 2.7K rows and 7 columns and contains the following features:

-   **Age**: Age of the individual.

-   **Sex**: Gender of the individual (male/female).

-   **BMI (Body Mass Index)**: A measure of body fat based on height and weight.

-   **Children**: Number of children/dependents covered by the insurance plan.

-   **Smoker**: Smoking status of the individual (yes/no).

-   **Region**: Geographical region in the US where the individual resides (northeast, southeast, southwest, northwest).

-   **Charges**: Medical insurance costs billed by the insurance company (target variable).

## Questions/Hypotheses to be addressed

*State the research questions you plan to answer with this analysis.*

The primary question we aim to answer using the "Medical Insurance Cost Prediction" dataset is:

**"What are the key factors influencing medical insurance costs, and how accurately can we predict these costs based on individuals' demographic and health-related attributes?"**

### Outcomes of Interest

The main outcome of interest in this analysis is:

-   **Medical Insurance Costs (charges)**: The amount billed by the insurance company for an individual's medical insurance.

### Specific Predictors

The specific predictors (independent variables) we will focus on include:

-   **Age**: How age influences insurance costs.

-   **Sex**: The impact of gender on insurance costs.

-   **BMI (Body Mass Index)**: The relationship between BMI and insurance costs.

-   **Children**: How the number of dependents affects insurance costs.

-   **Smoker**: The effect of smoking status on insurance costs.

-   **Region**: Regional differences in insurance costs within the US.

### Relations/Patterns of Interest

I am looking for several key relationships and patterns in the data, including:

1.  **Age and Insurance Costs**:Investigating whether older individuals tend to have higher medical insurance costs.

2.  **Gender Differences**:Analyzing whether there is a significant difference in insurance costs between males and females.

3.  **BMI and Insurance Costs**:Exploring the relationship between an individual's BMI and their medical insurance costs.Identifying any nonlinear patterns where very high or very low BMI values might correlate with higher insurance costs.

4.  **Impact of Smoking**:Assessing how smoking status (yes/no) affects insurance costs.Quantifying the cost difference between smokers and non-smokers.

5.  **Number of Children**: Examining whether having more dependents (children) increases insurance costs.

6.  **Regional Variations**: Identifying differences in insurance costs across different regions in the US (northeast, southeast, southwest, northwest).

# Methods

*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*

The methods used to clean, process and analyze the are as follows:

-   **Data Cleaning and Preprocessing:**
    -   Load the data
    -   Convert categorical variable
    -   Identify missing values
    -   Find Outliers
-   **Exploratory Data Analysis:**
    -   Summarize the data
    -   Visualize the data
-   **Model building and Analysis Approach:**
    -   Train-Test split the data
    -   Linear Regression Model
    -   Random Forest Model
    -   Model Evaluation

## Data import and cleaning

*Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along.*

The dataset used in this analysis is sourced from Kaggle. The dataset was downloaded directly from Kaggle website. It can be accessed at the following URL: [Medical Insurance Cost Prediction](https://www.kaggle.com/datasets/rahulvyasm/medical-insurance-cost-prediction).

The methods used to clean, process and analyze the are as follows:

-   **Data Cleaning and Preprocessing:**
    -   Load the data (can be found in processingcode2.R)
    -   Convert categorical variable (can be found in processingcode2.R)
    -   Identify missing values (can be found in processingcode2.R)
    -   Find Outlier (can be found in eda2.R)
-   **Exploratory Data Analysis:**
    -   Summarize the data (can be found in eda2.R)
    -   Visualize the data (can be found in eda2.R)
-   **Model building and Analysis Approach: *(Have not started yet)***
    -   Train-Test split the data
    -   Linear Regression Model
    -   Random Forest Model
    -   Model Evaluation

## Statistical analysis

The data will be built and analyzed as follows:

-   **Model building and Analysis Approach: *(Have not started yet)***
    -   Train-Test split the data
    -   Linear Regression Model
    -   Random Forest Model
    -   Model Evaluation

# Results

## Exploratory/Descriptive analysis

*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

@tbl-summarytable shows a summary of the data.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable2.rds")
knitr::kable(resulttable)
```

@fig-outliers-bmi shows the outliers in BMI

```{r}
#| label: fig-outliers-bmi
#| fig-cap: "BMI Outliers"
#| echo: FALSE
knitr::include_graphics(here("results","figures","bmi_outliers.png"))
```

@fig-outliers-charges shows the outliers in charges

```{r}
#| label: fig-outliers-charges
#| fig-cap: "Charges Outliers"
#| echo: FALSE
knitr::include_graphics(here("results","figures","charges-outliers.png"))

```

@fig-charges shows histogram of insurance charges

```{r}
#| label: fig-charges
#| fig-cap: "Distribution of Insurance charges"
#| echo: FALSE
knitr::include_graphics(here("results","figures","Charges-histogram.png"))
```

@fig-smoking shows box plot of insurance charges by smoking status

```{r}
#| label: fig-smoking
#| fig-cap: "Insurance Charges by smoking status."
#| echo: FALSE
knitr::include_graphics(here("results","figures","charges-smoking-boxplot.png"))

```

@fig-region shows box plot of insurance charges by region

```{r}
#| label: fig-region
#| fig-cap: "Insurance Charges by Region."
#| echo: FALSE
knitr::include_graphics(here("results","figures","charges-region-boxplot.png"))
```

@fig-age shows scatter plot of insurance charges by age

```{r}
#| label: fig-age
#| fig-cap: "Insurance Charges by Age."
#| echo: FALSE
knitr::include_graphics(here("results","figures","charge-age-boxplot.png"))
```

@fig-bmi shows scatter plot of insurance by BMI

```{r}
#| label: fig-bmi
#| fig-cap: "Insurance Charges by BMI."
#| echo: FALSE
knitr::include_graphics(here("results","figures","charge-bmi-scatter.png"))
```

@fig-children shows bar graph of average insurance charges by number of children

```{r}
#| label: fig-chidren
#| fig-cap: "Average Insurance Charges by Number of Children."
#| echo: FALSE
knitr::include_graphics(here("results","figures","children-bar.png"))
```

## Basic statistical analysis

*To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*

## Full analysis

*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here.*

```{}
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation

*Summarize what you did, what you found and what it means.*

## Strengths and Limitations

*Discuss what you perceive as strengths and limitations of your analysis.*

## Conclusions

*What are the main take-home messages?*

*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

{{< pagebreak >}}

# References

### 2024-06-12 code/processing-code/processingcode2.R

@@ -1,0 +1,78 @@
+###############################
+# processing script
+#
+#this script loads the raw data, processes and cleans it 
+#and saves it as Rds file in the processed-data folder
+#
+# Note the ## ---- name ---- notation
+# This is done so one can pull in the chunks of code into the Quarto document
+# see here: https://bookdown.org/yihui/rmarkdown-cookbook/read-chunk.html
+###############################
+
+
+## ---- packages --------
+#load needed packages. make sure they are installed.
+library(readxl) #for loading Excel files
+library(dplyr) #for data processing/cleaning
+library(tidyr) #for data processing/cleaning
+library(skimr) #for nice visualization of data 
+library(here) #to set paths
+
+
+## ---- loaddata --------
+#path to data
+#note the use of the here() package and not absolute paths
+data_location <- here::here("data","raw-data","medical_insurance2.csv")
+#load data. 
+#note that for functions that come from specific packages (instead of base R)
+# I often specify both package and function like so
+#package::function() that's not required one could just call the function
+#specifying the package makes it clearer where the function "lives",
+#but it adds typing. You can do it either way.
+rawdata <- read.csv(data_location)
+
+
+## ---- exploredata --------
+#take a look at the data
+dplyr::glimpse(rawdata)
+#another way to look at the data
+summary(rawdata)
+#yet another way to get an idea of the data
+head(rawdata)
+#this is a nice way to look at data
+skimr::skim(rawdata)
+
+# Convert categorical variables to factors
+rawdata$sex <- as.factor(rawdata$sex)
+rawdata$smoker <- as.factor(rawdata$smoker)
+rawdata$region <- as.factor(rawdata$region)
+
+## ---- savedata --------
+# all done, data is clean now. 
+# Let's assign at the end to some final variable
+# makes it easier to add steps above
+processeddata <- rawdata
+# location to save file
+save_data_location <- here::here("data","processed-data","processeddata2.rds")
+saveRDS(processeddata, file = save_data_location)
+
+
+
+## ---- notes --------
+# anything you don't want loaded into the Quarto file but 
+# keep in the R file, just give it its own label and then don't include that label
+# in the Quarto file
+
+# Dealing with NA or "bad" data:
+# removing anyone who had "faulty" or missing data is one approach.
+# it's often not the best. based on your question and your analysis approach,
+# you might want to do cleaning differently (e.g. keep individuals with some missing information)
+
+# Saving data as RDS:
[…]

### 2024-06-18 code/eda-code/eda2.qmd
commit changes

@@
 
-Once more height as function of weight, stratified by gender. Note that there is so little data, it's a bit silly. But we'll plot it anyway.
+Now let's look at a boxplot for charges by smoking status
 
@@
 
+Boxplot for charges by region
+
 ```{r}
@@
 
+Scatter plot for charges by age
 ```{r}
@@
 
+Scatter plot for charges by BMI
 ```{r}
@@
 
+Bar Plot if average charges by number of children
 ```{r}

### 2024-06-18 products/manuscript/Manuscript.qmd
commit changes

@@
 
+@tbl-resulttable1 shows a simple model with one predictor using linear regression analysis.
+
+```{r}
+#| label: tbl-resulttable1
+#| tbl-cap: "Simple Models with One Predictor."
+#| echo: FALSE
+simpletable=readRDS("../../results/tables/simple_models_results.rds")
+knitr::kable(simpletable)
+```
+
 ## Full analysis
@@
 
-```{}
+@tbl-resulttable2 shows linear regression analysis charges is analyzed across all predictors.
+
+```{r}
+#| label: tbl-resulttable2
+#| tbl-cap: "Linear Regression Analysis."
+#| echo: FALSE
+lineartable=readRDS("../../results/tables/linear_regression_results.rds")
+knitr::kable(lineartable)
+
+```
+
+@fig-linear helps visualize which features have a significant effect on charges.
+
+```{r}
+#| label: fig-linear
+#| fig-cap: "Linear Regression Coefficients."
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","linear_regression_plot.png"))
+
+```
+
+@tbl-resulttable3 helps to capture complex interaction between features.
+
+```{r}
+#| label: tbl-resulttable3
+#| tbl-cap: "Random Forest Analysis."
+#| echo: FALSE
+rf_table=readRDS("../../results/tables/random_forest_importance.rds")
+knitr::kable(rf_table)
+```
+
+@fig-randomforest helps visualize how each feature have a significant effect on charges.
+
+```{r}
+#| label: fig-randomforest
+#| fig-cap: "Random Forest Feature Importance."
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","random_forest_importance_plot.png"))
 ```

### 2024-06-18 code/analysis-code/statistical-analysis2.R

@@ -1,0 +1,95 @@
+###############################
+# analysis script
+#
+#this script loads the processed, cleaned data, does a simple analysis
+#and saves the results to the results folder
+
+#load needed packages. make sure they are installed.
+library(ggplot2) #for plotting
+library(broom) #for cleaning up output from lm()
+library(here) #for data loading/saving
+library(randomForest) #for random forest analysis
+
+#path to data
+#note the use of the here() package and not absolute paths
+data_location <- here::here("data","processed-data","processeddata2.rds")
+
+#load data. 
+mydata <- readRDS(data_location)
+
+
+######################################
+#Data fitting/statistical analysis
+######################################
+
+######################################
+# Simple Models with One Predictor
+######################################
+
+# List of predictor variables
+predictors <- c("age", "bmi", "children", "sex", "smoker", "region")
+
+# Initialize a list to store results
+results_list <- list()
+
+# Loop through each predictor and fit a simple linear regression model
+for (predictor in predictors) {
+  formula <- as.formula(paste("charges ~", predictor))
+  lmfit <- lm(formula, data = mydata)
+  
+  # Place results from fit into a data frame with the tidy function
+  lmtable <- broom::tidy(lmfit)
+  
+  # Add the results to the list
+  results_list[[predictor]] <- lmtable
+  
+  # Print the results for each model
+  print(paste("Results for predictor:", predictor))
+  print(lmtable)
+}
+
+# Combine all results into a single data frame
+results_df <- do.call(rbind, lapply(names(results_list), function(x) {
+  cbind(predictor = x, results_list[[x]])
+}))
+
+# Save the combined results table
+table_file <- here("results", "tables", "simple_models_results.rds")
+saveRDS(results_df, file = table_file)
+
+# Print combined results
+print(results_df)
+
+############################
+#### Linear regression analysis
+# fit linear model using charges as outcome, age, bmi, children, sex, smoker and region as predictors
+
+lmfit1 <- lm(charges ~ age + bmi + children + sex + smoker + region, mydata)
+
+
+# place results from fit into a data frame with the tidy function
+lmtable1 <- broom::tidy(lmfit1)
+
+#look at fit results
+print(lmtable1)
+
+# save fit results table  
+table_file1 = here("results", "tables", "linear_regression_results.rds")
+saveRDS(lmtable1, file = table_file1)
+
+############################
+#### Random Forest Analysis
[…]

### 2024-06-18 code/analysis-code/statistical-analysis3.R

@@ -1,0 +1,32 @@
+# This script creates visualizations for the analysis results.
+
+# Load needed packages
+library(ggplot2)
+library(here)
+
+# Load linear regression results
+linear_regression_results <- readRDS(here::here("results", "tables", "linear_regression_results.rds"))
+
+# Load random forest importance results
+random_forest_importance <- readRDS(here::here("results", "tables", "random_forest_importance.rds"))
+
+# Create a plot for linear regression coefficients
+plot_lm <- ggplot(linear_regression_results, aes(x = term, y = estimate)) +
+  geom_bar(stat = "identity") +
+  coord_flip() +
+  theme_minimal() +
+  labs(title = "Linear Regression Coefficients", x = "Predictor", y = "Estimate")
+
+# Save the linear regression plot
+ggsave(here("results", "figures", "linear_regression_plot.png"), plot = plot_lm)
+
+# Create a plot for random forest feature importance
+plot_rf <- ggplot(random_forest_importance, aes(x = reorder(Feature, Importance), y = Importance)) +
+  geom_bar(stat = "identity") +
+  coord_flip() +
+  theme_minimal() +
+  labs(title = "Random Forest Feature Importance", x = "Feature", y = "Importance") 
+
+
+# Save the random forest plot
+ggsave(here("results", "figures", "random_forest_importance_plot.png"), plot = plot_rf)

### 2024-06-18 products/manuscript/Manuscript.qmd
Commit changes

@@
 
-*Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section.*
-
 The dataset used for this analysis is sourced from Kaggle. The dataset comprises 2.7K rows and 7 columns and contains the following features:
@@
 ## Questions/Hypotheses to be addressed
-
-*State the research questions you plan to answer with this analysis.*
 
@@
 
-*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*
-
 The methods used to clean, process and analyze the are as follows:
@@
 ## Data import and cleaning
-
-*Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along.*
 
@@
 ## Exploratory/Descriptive analysis
-
-*Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project.*
 
@@
 
-*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
-
 @tbl-resulttable1 shows a simple model with one predictor using linear regression analysis.
@@
 ## Full analysis
-
-*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here.*
 

### 2024-06-18 code/analysis-code/statistical-analysis2.R

@@ -1,5 +1,5 @@
 ###############################
-# analysis script
+# analysis script for simple model with one predictor
 #
 #this script loads the processed, cleaned data, does a simple analysis
 #and saves the results to the results folder
@@ -8,7 +8,6 @@
 library(ggplot2) #for plotting
 library(broom) #for cleaning up output from lm()
 library(here) #for data loading/saving
-library(randomForest) #for random forest analysis
 
 #path to data
 #note the use of the here() package and not absolute paths
@@ -59,37 +58,3 @@ saveRDS(results_df, file = table_file)
 
 # Print combined results
 print(results_df)
-
-############################
-#### Linear regression analysis
-# fit linear model using charges as outcome, age, bmi, children, sex, smoker and region as predictors
-
-lmfit1 <- lm(charges ~ age + bmi + children + sex + smoker + region, mydata)
-
-
-# place results from fit into a data frame with the tidy function
-lmtable1 <- broom::tidy(lmfit1)
-
-#look at fit results
-print(lmtable1)
-
-# save fit results table  
-table_file1 = here("results", "tables", "linear_regression_results.rds")
-saveRDS(lmtable1, file = table_file1)
-
-############################
-#### Random Forest Analysis
-# let's capture complex interaction between features
-
-model_rf <- randomForest(charges ~ ., mydata, ntree = 100)  
-
-# Extract feature importance
-importance_rf <- importance(model_rf)
-importance_df <- data.frame(Feature = rownames(importance_rf), Importance = importance_rf[, 1])
-
-# look at feature importance
-print(importance_df)
-
-# save feature importance table  
-table_file2 <- here("results", "tables", "random_forest_importance.rds")
-saveRDS(importance_df, file = table_file2)

### 2024-06-18 code/analysis-code/statistical-analysis4.R

@@ -1,0 +1,94 @@
+###############################
+# analysis script to perform comprehensive analysis involve training and testing the data
+# running linear regression and random forest models and 
+# finally conducting model evaluation
+
+#this script loads the processed, cleaned data, does a simple analysis
+#and saves the results to the results folder
+
+# Load needed packages
+library(ggplot2) # for plotting
+library(broom) # for cleaning up output from lm()
+library(here) # for data loading/saving
+library(randomForest) # for random forest analysis
+library(caret) # for train-test split and model evaluation
+
+# Path to data
+data_location <- here::here("data","processed-data","processeddata2.rds")
+
+# Load data
+mydata <- readRDS(data_location)
+
+# Split the data into training and testing sets
+set.seed(123) # for reproducibility
+trainIndex <- createDataPartition(mydata$charges, p = 0.8, list = FALSE)
+data_train <- mydata[trainIndex, ]
+data_test <- mydata[-trainIndex, ]
+
+# Fit Linear Regression Model
+lmfit <- lm(charges ~ age + bmi + children + sex + smoker + region, data = data_train)
+
+# Predict on the test set
+lm_predictions <- predict(lmfit, newdata = data_test)
+
+# Evaluate Linear Regression Model
+lm_rmse <- sqrt(mean((data_test$charges - lm_predictions)^2))
+lm_mae <- mean(abs(data_test$charges - lm_predictions))
+lm_r2 <- 1 - sum((data_test$charges - lm_predictions)^2) / sum((data_test$charges - mean(data_test$charges))^2)
+
+cat("Linear Regression Model Evaluation:\n")
+cat("RMSE: ", lm_rmse, "\n")
+cat("MAE: ", lm_mae, "\n")
+cat("R-squared: ", lm_r2, "\n")
+
+# Save Linear Regression Results
+lmtable <- broom::tidy(lmfit)
+table_file1 <- here("results", "tables", "linear_regression_results.rds")
+saveRDS(lmtable, file = table_file1)
+
+# Create a plot for linear regression coefficients
+plot_lm <- ggplot(lmtable, aes(x = reorder(term, estimate), y = estimate, fill = term)) +
+  geom_bar(stat = "identity") +
+  coord_flip() +
+  theme_minimal() +
+  labs(title = "Linear Regression Coefficients", x = "Term", y = "Estimate") +
+  scale_fill_brewer(palette = "Set2")
+
[…]

### 2024-06-18 products/manuscript/Manuscript.qmd
commit changes

@@
 ```{r}
-#| label: fig-chidren
+#| label: fig-children
 #| fig-cap: "Average Insurance Charges by Number of Children."

### 2024-07-16 products/manuscript/Manuscript.qmd
commit changes

@@
 
-@fig-outliers-charges shows the outliers in charges
+Observation: There are several outliers with BMI values above 50, indicating that there are individuals with unusually high BMI compared to the rest of the dataset.\@fig-outliers-charges shows the outliers in charges
 
@@
 
+Observation: There are numerous outliers with insurance charges above 40,000, indicating that there are individuals with unusually high insurance charges compared to the rest of the dataset.
+
 @fig-charges shows histogram of insurance charges
@@
 ```
+
+Observation: The histogram shows a right-skewed distribution, indicating that most individuals have lower insurance charges, with fewer individuals having very high charges.
 
@@
 
+Observation: There is a clear and substantial difference in insurance charges between smokers and non-smokers.
+Smokers have much higher charges on average.
+Charges for smokers are more variable than for non-smokers, with a wider range of values.
+High charges are more common among smokers and are not considered outliers, whereas for non-smokers, such high charges are less common and considered outliers.
+
 @fig-region shows box plot of insurance charges by region
@@
 ```
+
+Observation: The Southeast region has the highest median insurance charges, followed by the Northeast, Northwest, and Southwest.
+The Southwest region has the lowest median charges.
+All regions have numerous outliers with very high charges, particularly the Southeast and Northeast.
 
@@
 
+Observation: There is a clear upward trend in insurance charges as age increases.
+This suggests that insurance charges tend to increase with age.
+The relationship appears to be somewhat linear, with higher age associated with higher charges, but there is a noticeable spread in the data.
+
 @fig-bmi shows scatter plot of insurance by BMI
@@
 
+Observation: There is a noticeable upward trend in insurance charges as BMI increases.
+This suggests that individuals with higher BMI tend to have higher insurance charges.
+The relationship appears to be somewhat nonlinear, with higher charges associated with higher BMI values, but there is considerable spread in the data.
+
 @fig-children shows bar graph of average insurance charges by number of children
@@
 ```
+
+Observation: There is a general trend where insurance charges increase with the number of children up to three children, after which there is a slight decrease for four children, and a significant decrease for five children.
+The drop in average charges for households with five children could be due to various reasons such as discounts for larger families, different insurance plans, or data anomalies.
 
@@
 
+Observation:
+
+-   Age, BMI, number of children, sex, smoking status, and living in the Southeast and Southwest regions are significant predictors of insurance charges.
+
+-   Smoking status has the largest effect on insurance charges, significantly increasing them.
+
+-   The intercept for BMI and living in the Northwest region are not significant predictors of insurance charges in these models.
+
 ## Full analysis
@@
 ```
+
+Observation: After the data was split into training and testing, the linear regression analysis was conducted.
+
+-   Age, BMI, number of children, smoking status, and living in the Southeast or Southwest regions are significant predictors of insurance charges.
+
+-   Smoking status still has the largest effect on insurance charges, significantly increasing them.
+
+-   Sex and living in the Northwest region are not significant predictors of insurance charges in this model.
 
@@
 
+Observation:
+
+-   Smoking status, BMI, and age are the most important predictors of insurance charges, followed by the number of children and region. Sex has the least importance.
+
+-   The importance rankings from the random forest analysis are consistent with the findings from the linear regression models, where smoking status, age, and BMI were significant predictors.
+
 @fig-randomforest helps visualize how each feature have a significant effect on charges.
@@
 
+### What was done:
+
+-   **Data Loading and Preparation**:
+
+    -   Loaded the insurance dataset from a CSV file.
+
+    -   Performed exploratory data analysis to understand the distribution and relationships within the data.
+
+    -   Conducted data cleaning processes, such as handling missing values and transforming variables.
+
+-   **Exploratory Data Analysis (EDA)**:
+
+    -   Visualized the distribution of insurance charges using histograms and box plots.
+
+    -   Examined the relationships between insurance charges and various predictors, including age, BMI, number of children, smoking status, and region.
+
+-   **Statistical Modeling**:
+
+    -   Built simple linear regression models to explore the impact of individual predictors on insurance charges.
+
+    -   Developed a multiple linear regression model to understand the combined effect of multiple predictors on insurance charges.
+
+    -   Conducted a random forest analysis to determine the importance of different features in predicting insurance charges.
+
+-   **Visualization**:
+
+    -   Created various plots, including scatter plots, bar charts, and box plots, to visualize the relationships between predictors and insurance charges.
+
+### What was found:
+
+-   **Key Predictors**:
+
+    -   **Smoking Status**: The most influential predictor, with smokers incurring significantly higher insurance charges compared to non-smokers.
+
+    -   **BMI**: A strong positive relationship with insurance charges, indicating higher charges for individuals with higher BMI.
+
+    -   **Age**: A positive relationship with insurance charges, with charges increasing as age increases.
+
+    -   **Number of Children**: A significant positive impact on insurance charges, although the effect was smaller compared to smoking status, BMI, and age.
+
+    -   **Region**: Regional differences were observed, with the Southeast region associated with higher charges and the Southwest region associated with lower charges.
+
+    -   **Sex**: The impact of sex on insurance charges was negligible.
+
+-   **Feature Importance**:
+
+    -   The random forest analysis confirmed that smoking status, BMI, and age were the most important features in predicting insurance charges.
+
+    -   The number of children and region also contributed to the predictions, while sex had the least importance.
+
+### What it means:
+
+-   **Policy and Pricing**:
+
+    -   The findings highlight the importance of considering multiple factors when setting insurance premiums.
+Smoking status, BMI, and age are critical factors that significantly impact insurance costs.
+
+    -   Insurance companies can use these insights to adjust premiums more accurately based on the risk profile of individuals.
+
+-   **Health Interventions**:
+
+    -   The strong association between smoking and higher insurance charges underscores the need for targeted health interventions to reduce smoking rates, which could, in turn, lower insurance costs.
+
+    -   Addressing obesity through health programs could also help reduce insurance charges associated with higher BMI.
+
+-   **Regional Considerations**:
+
+    -   Regional differences in insurance charges suggest that localized strategies might be necessary to address specific regional health challenges and cost structures.
+
+-   **Further Research**:
+
+    -   Future studies could explore the interactions between these key predictors to understand their combined effects on insurance charges.
+
+    -   Additional data could be collected to validate these findings and explore other potential predictors.
+
 ## Strengths and Limitations
@@
 
+### Strengths:
+
+1.  **Comprehensive Exploratory Data Analysis (EDA)**:
+
+    -   Detailed EDA helped in understanding the distribution and relationships within the data.
+
+    -   Visualizations such as histograms, box plots, scatter plots, and bar charts provided clear insights into the data and highlighted important patterns.
+
+2.  **Use of Multiple Statistical Methods**:
+
+    -   Employed both linear regression and random forest models to analyze the data.
+
+    -   Linear regression provided insights into the individual and combined effects of predictors.
+
+    -   Random forest analysis helped in identifying the importance of different features.
+
+3.  **Handling of Categorical Variables**:
+
+    -   Effectively transformed and factored categorical variables to ensure accurate modeling.
+
+    -   Considered regional differences and the impact of smoking status, which are critical factors in insurance pricing.
+
+4.  **Feature Importance Analysis**:
+
+    -   The random forest feature importance analysis provided a clear ranking of predictors, confirming the results of the linear regression models.
+
+5.  **Clear and Informative Visualizations**:
+
+    -   Created visualizations that were easy to interpret and clearly illustrated the relationships between variables and insurance charges.
+
+6.  **Thorough Interpretation**:
+
+    -   Provided detailed interpretations of the results, explaining the significance and impact of each predictor on insurance charges.
+
+### Limitations:
+
+-   **Assumptions of Linear Regression**:
+
+    -   Linear regression models assume linear relationships between predictors and the outcome, which may not fully capture more complex, non-linear relationships.
+
+    -   Assumes independence of observations, which might not hold true in real-world data.
+
+-   **Potential Overfitting with Random Forest**:
+
+    -   Random forest models can be prone to overfitting, especially with a small dataset or highly correlated features.
+
+    -   Overfitting might lead to overly optimistic feature importance scores that do not generalize well to new data.
+
+-   **Outliers and Data Distribution**:
+
+    -   The dataset contains outliers, particularly in insurance charges, which can impact the results of linear regression models.
+
+    -   The skewed distribution of insurance charges might affect the assumptions of normality in linear regression.
+
+-   **Limited Dataset**:
+
+    -   The analysis was conducted on a specific insurance dataset, which may limit the generalizability of the findings to other datasets or populations.
+
+    -   Potential biases in the dataset could affect the results and their applicability.
+
 ## Conclusions
@@
 *What are the main take-home messages?*
+
+The analysis highlights the significant factors affecting insurance charges and provides valuable insights for insurance pricing and policy formulation.
+By focusing on key predictors such as smoking status, BMI, age, number of children, and region, and considering their combined effects, insurance companies can develop more accurate and equitable pricing strategies.
+Additionally, public health interventions targeting smoking and obesity can help reduce overall insurance costs.
+Continued research and validation are essential to further refine these insights and ensure their applicability across diverse populations and datasets.
 

### 2024-07-28 products/manuscript/Manuscript.qmd
commit changes

@@
 
+{{< pagebreak >}}
+
 # Results
@@
 
+{{< pagebreak >}}
+
 @fig-outliers-bmi shows the outliers in BMI
@@
 
-Observation: There are several outliers with BMI values above 50, indicating that there are individuals with unusually high BMI compared to the rest of the dataset.\@fig-outliers-charges shows the outliers in charges
+Observation: There are several outliers with BMI values above 50, indicating that there are individuals with unusually high BMI compared to the rest of the dataset.\@fig-outliers-charges shows the outliers in charges.
+
+{{< pagebreak >}}
+
+@fig-outliers-charges shows the outliers in Charges
 
@@
 
+{{< pagebreak >}}
+
 @fig-charges shows histogram of insurance charges
@@
 Observation: The histogram shows a right-skewed distribution, indicating that most individuals have lower insurance charges, with fewer individuals having very high charges.
+
+{{< pagebreak >}}
 
@@
 
+{{< pagebreak >}}
+
 @fig-region shows box plot of insurance charges by region
@@
 All regions have numerous outliers with very high charges, particularly the Southeast and Northeast.
+
+{{< pagebreak >}}
 
@@
 
+{{< pagebreak >}}
+
 @fig-bmi shows scatter plot of insurance by BMI
@@
 
+{{< pagebreak >}}
+
 @fig-children shows bar graph of average insurance charges by number of children
@@
 The drop in average charges for households with five children could be due to various reasons such as discounts for larger families, different insurance plans, or data anomalies.
+
+{{< pagebreak >}}
 
@@
 -   The intercept for BMI and living in the Northwest region are not significant predictors of insurance charges in these models.
+
+{{< pagebreak >}}
 
@@
 
+{{< pagebreak >}}
+
 @fig-linear helps visualize which features have a significant effect on charges.
@@
 ```
+
+{{< pagebreak >}}
 
@@
 -   The importance rankings from the random forest analysis are consistent with the findings from the linear regression models, where smoking status, age, and BMI were significant predictors.
+
+{{< pagebreak >}}
 
@@
 
-*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
+This website [@careinsurance_smoker] discusses the challenges smokers face in obtaining health insurance, highlighting how smoking affects premium costs..
 
-This paper [@leek2015] discusses types of analyses.
+This Experian website [@experian_premium_factors] explains various factors that affect health insurance premium costs, including age, location, tobacco use, type of plan, and individual vs.
+family enrollment.
+It provides insights on how these factors influence the overall cost of health insurance for consumers..
 
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like.
+The article [@choi_blackburn_2018] examines how household sociodemographic characteristics influence health insurance premiums and medical expenses.
+Using data from the 2014 Consumer Expenditure Survey, the study identifies that age, marital status, educational attainment, and family income are significant predictors of higher spending on health insurance premiums and medical expenses.
+.
 

### 2024-08-05 products/manuscript/Manuscript.qmd
commit changes

@@
 
-Observation: There are several outliers with BMI values above 50, indicating that there are individuals with unusually high BMI compared to the rest of the dataset.\@fig-outliers-charges shows the outliers in charges.
+**Observation:** There are several outliers with BMI values above 50, indicating that there are individuals with unusually high BMI compared to the rest of the dataset.\@fig-outliers-charges shows the outliers in charges.
 
@@
 
-Observation: There are numerous outliers with insurance charges above 40,000, indicating that there are individuals with unusually high insurance charges compared to the rest of the dataset.
+**Observation:** There are numerous outliers with insurance charges above 40,000, indicating that there are individuals with unusually high insurance charges compared to the rest of the dataset.
 
@@
 
-Observation: The histogram shows a right-skewed distribution, indicating that most individuals have lower insurance charges, with fewer individuals having very high charges.
+**Observation:** The histogram shows a right-skewed distribution, indicating that most individuals have lower insurance charges, with fewer individuals having very high charges.
 
@@
 
-Observation: There is a clear and substantial difference in insurance charges between smokers and non-smokers.
+**Observation:** There is a clear and substantial difference in insurance charges between smokers and non-smokers.
 Smokers have much higher charges on average.
@@
 
-Observation: The Southeast region has the highest median insurance charges, followed by the Northeast, Northwest, and Southwest.
+**Observation:** The Southeast region has the highest median insurance charges, followed by the Northeast, Northwest, and Southwest.
 The Southwest region has the lowest median charges.
@@
 
-Observation: There is a clear upward trend in insurance charges as age increases.
+**Observation:** There is a clear upward trend in insurance charges as age increases.
 This suggests that insurance charges tend to increase with age.
@@
 
-Observation: There is a noticeable upward trend in insurance charges as BMI increases.
+**Observation:** There is a noticeable upward trend in insurance charges as BMI increases.
 This suggests that individuals with higher BMI tend to have higher insurance charges.
@@
 
-Observation: There is a general trend where insurance charges increase with the number of children up to three children, after which there is a slight decrease for four children, and a significant decrease for five children.
+**Observation:** There is a general trend where insurance charges increase with the number of children up to three children, after which there is a slight decrease for four children, and a significant decrease for five children.
 The drop in average charges for households with five children could be due to various reasons such as discounts for larger families, different insurance plans, or data anomalies.
@@
 
-Observation:
+**Observation:**
 
@@
 
-Observation: After the data was split into training and testing, the linear regression analysis was conducted.
+**Observation:** After the data was split into training and testing, the linear regression analysis was conducted.
 
@@
 ## Summary and Interpretation
-
-*Summarize what you did, what you found and what it means.*
 
@@
 
-*Discuss what you perceive as strengths and limitations of your analysis.*
-
 ### Strengths:
@@
 
-*What are the main take-home messages?*
-
 The analysis highlights the significant factors affecting insurance charges and provides valuable insights for insurance pricing and policy formulation.
@@
 Continued research and validation are essential to further refine these insights and ensure their applicability across diverse populations and datasets.
+
+A few online resources that helped with my research:
 

### 2024-08-05 code/processing-code/processingcode2.R

@@ -57,22 +57,3 @@ save_data_location <- here::here("data","processed-data","processeddata2.rds")
 saveRDS(processeddata, file = save_data_location)
 
 
-
-## ---- notes --------
-# anything you don't want loaded into the Quarto file but 
-# keep in the R file, just give it its own label and then don't include that label
-# in the Quarto file
-
-# Dealing with NA or "bad" data:
-# removing anyone who had "faulty" or missing data is one approach.
-# it's often not the best. based on your question and your analysis approach,
-# you might want to do cleaning differently (e.g. keep individuals with some missing information)
-
-# Saving data as RDS:
-# I suggest you save your processed and cleaned data as RDS or RDA/Rdata files. 
-# This preserves coding like factors, characters, numeric, etc. 
-# If you save as CSV, that information would get lost.
-# However, CSV is better for sharing with others since it's plain text. 
-# If you do CSV, you might want to write down somewhere what each variable is.
-# See here for some suggestions on how to store your processed data:
-# http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

### 2024-08-05 products/manuscript/Manuscript.qmd
commit changes

@@
 bibliography: ../../assets/dataanalysis-references.bib
-csl: ../../assets/apa.csl
 ---

### 2024-08-06 code/eda-code/eda2.qmd
commit changes

@@
 This Quarto file loads the cleaned data and does some exploring.
-
-I'm only showing it the way where the code is included in the file. 
-As described in the `processing_code` materials, I currently prefer the approach of having R code in a separate file and pulling it in.
-
-But I already had this written and haven't yet re-done it that way. Feel free to redo and send a pull request on GitHub :)
-
-Again, it is largely a matter of preference and what makes the most sense to decide if one wants to have code inside Quarto files, or as separate R files.
-And sometimes, an R script with enough comments is good enough and one doesn't need a Quarto file.
-
-Also note that while here I split cleaning and exploring, this is iterative.
-You saw that as part of the processing, we already had to explore the data somewhat to understand how to clean it.
-In general, as you explore, you'll find things that need cleaning.
-As you clean, you can explore more.
-Therefore, at times it might make more sense to combine the cleaning and exploring code parts into a single R or Quarto file.
-Or split things in any other logical way.
-
-As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data.
-Depending on the total number of variables in your dataset, explore all or some of the others.
-Figures produced here might be histograms or density plots, correlation plots, etc.
-Tables might summarize your data.
-
-Start by exploring one variable at a time.
-Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in.
-If your dataset is small, you can do that for all variables.
-
-Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.
 
@@
 And/or you could have separate folders for exploratory tables/figures and for final tables/figures.
-Just choose a setup that makes sense for your project and works for you, and provide enough documentation that someone can understand what you are doing.
 
@@
 ```
-
-# Notes
-
-For your own explorations, tables and figures can be "quick and dirty".
-As long as you can see what's going on, there is no need to polish them.
-That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.).
-Those should look as nice, polished and easy to understand as possible.

### 2024-08-06 code/analysis-code/statistical-analysis4.R

@@ -3,7 +3,8 @@
 # running linear regression and random forest models and 
 # finally conducting model evaluation
 
-#this script loads the processed, cleaned data, does a simple analysis
+#this script loads the processed, cleaned data, does a linear regression and
+# random forest analysis
 #and saves the results to the results folder
 
 # Load needed packages

### 2024-08-06 code/processing-code/processingcode2.R

@@ -3,10 +3,7 @@
 #
 #this script loads the raw data, processes and cleans it 
 #and saves it as Rds file in the processed-data folder
-#
-# Note the ## ---- name ---- notation
-# This is done so one can pull in the chunks of code into the Quarto document
-# see here: https://bookdown.org/yihui/rmarkdown-cookbook/read-chunk.html
+# 
 ###############################
 
 
